---
title: "Exercises_W2"
author: "Brihatee Mattapullut"
date: 06/10/2026 # Automatically uses today's date
date-format: "DD/MM/YYYY"             # British/Australian date format

# Output Config
format: 
  # HTML Output
  html:
    theme: cosmo          # Modern Bootstrap theme with clean typography
    highlight-style: a11y # High-contrast syntax highlighting for code
    toc: true             # Includes a Table of Contents
    toc-depth: 3          # Captures down to sub-subsections (e.g., 3.1.1)
    toc-location: left    # Pins TOC to left sidebar so it doesn't block main page
    toc-title: "Contents" # Custom title for the TOC sidebar
    number-sections: true # Numbers sections automatically (e.g., 1, 1.1, 2)
    code-fold: true       # Collapses code blocks by default under a dropdown
    code-summary: "Show code" # Label text on collapsed code blocks
    code-tools: true      # Adds global code menu (Show/Hide All Code) at top right
    code-copy: true       # Adds a one-click copy button to code blocks
    embed-resources: true # Bundles images/CSS into a single stand-alone HTML file
    df-print: paged       # Formats data frames into interactive paged tables
    fig-height: 5         # Default plot height in inches
    fig-width: 8          # Default plot width in inches
    #fig-responsive: true # Scales figures smoothly across screen sizes
    smooth-scroll: true   # Enables smooth scrolling when clicking TOC links
---

# Loading libraries
```{r}
library(tidyverse)
library(readxl)
```


# 2.12 Keystone exercise: Estuary Fish Survey - Data Rescue Mission


# Importing the files
```{r}

#estuary catch log data
#the excel workbook has different tabs 
#need to be loaded individually
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)  

#estuary metadata file
metadata <- read.csv(here::here("data/workshop2/estuary_metadata.csv")) 
  
#estuary sonde data file
# code is written in separate script

source(here::here("code/diw_sensor.R"))
  
#species dictionary file
species_dict <- 
read.csv(here::here("data/workshop2/species_dictionary.csv"))
```


# Inspecting the indiviual tabs

Look at anomalies between the workbooks and variable names if they do not align

The column titles are consistent but the variable names are not consistent across the different tabs.
```{r inspect-catch-tabs}
catch1
catch2
catch3
catch4

```


# Combining all the different tabs in the catch log excel 

Binding the rows

```{r bind-catch}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```


# Preliminary checks

```{r inspect-catch}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)

```


Using distinct() to check for unique values in the columns to see if there are any anomalies in the data and if need to clean before converting to factor variables
  
    - the species names and sites are not consistent across the different tabs. The species and site names have different capitalisation. 
    - need to be cleaned before joining with the species dictionary and estuary metadata files.
    
    
```{r distinct-catch}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)

```

# Checking metadata file

    - Site names also need to be cleaned before joining with the catch data file.
    
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```



# Fixing site names in the catch data file + converting to factor variables

```{r fix-site-names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

```

- checking if the site names have been fixed

```{r recheck-site-names}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
str(metadata)

```

# Checking taxonomic dictionary file

Common names need to be checked in the catch log and the species dictionary file to see if they are consistent before joining the two files.

```{r}
#checking species dictionary file
head(species_dict)
```
```{r check-species-names}
catch_data |> 
  distinct(species)
# Can see these are different to the dictionary
species_dict |>
  distinct(common_name)

```
## Standardising species names in the catch data file to match the species dictionary file

Even after standardising, the species name still is not consistent with the species dictionary file.


```{r fix-species-names}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

## Creating new variable in catch log file to match the species dictionary file

after creating the new variable, the species names need to be joined with the species dictionary file to get the correct species names.

```{r join-species-dict}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))

```

```{r}
# checking the catch log to see if the species and site names are fixed

glimpse(catch_data)
```

## Sensor data

The sensor data was taken for a duration of 15 minutes for each four sites for 30 days. The data was collected at 1 second intervals. The sensor data was collected using a YSI EXO2 sonde. 

The data was collected for the following parameters: temperature, salinity, dissolved oxygen, pH, turbidity, and chlorophyll-a.

The data was collected for the following sites: Lower Ross, Upper Ross, Lower Pioneer, Upper Pioneer.


## Joining the sensor data with the metadata file

```{r join-all}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))

```

```{r check-final-join}
glimpse(catch_data_final)
#checking if the join worked
```

There should be 480 rows of data for each site (4 sites x 30 days x 4 parameters) and 12 columns of data (catch log + metadata + sensor data). But we only have 427 rows of data. This is because there are some missing values in the sensor data file. The missing values are due to the sonde recording zero counts of certain species at different sites. 
## Wrangling final dataset

```{r wrangle-final-zeros}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )

```

After the main wrangling, need to check the difference in environmental variables among the zones.

### Salinity
```{r}
# Let's make a table
sal_sum <- 
  catch_data_final |>
  group_by(site) |>
  summarise(
   # sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# And from this a plot
sal_sum |> 
  ggplot(aes(x = site, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Site",
       y = "Median Salinity (ppt)")

```







