---
title: "Module 2 Workshop 2"
author: "Brihatee Mattapullut"
date: 30/09/2026 # Automatically uses today's date
date-format: "DD/MM/YYYY"             # British/Australian date format

# Output Config
format: 
  # HTML Output
  html:
    theme: cosmo          # Modern Bootstrap theme with clean typography
    highlight-style: a11y # High-contrast syntax highlighting for code
    toc: true             # Includes a Table of Contents
    toc-depth: 3          # Captures down to sub-subsections (e.g., 3.1.1)
    toc-location: left    # Pins TOC to left sidebar so it doesn't block main page
    toc-title: "Contents" # Custom title for the TOC sidebar
    number-sections: true # Numbers sections automatically (e.g., 1, 1.1, 2)
    code-fold: true       # Collapses code blocks by default under a dropdown
    code-summary: "Show code" # Label text on collapsed code blocks
    code-tools: true      # Adds global code menu (Show/Hide All Code) at top right
    code-copy: true       # Adds a one-click copy button to code blocks
    embed-resources: true # Bundles images/CSS into a single stand-alone HTML file
    df-print: paged       # Formats data frames into interactive paged tables
    fig-height: 5         # Default plot height in inches
    fig-width: 8          # Default plot width in inches
    #fig-responsive: true # Scales figures smoothly across screen sizes
    smooth-scroll: true   # Enables smooth scrolling when clicking TOC links
---

# 2.2 Tidying data - Loading libraries
```{r}
library(tidyverse)
```


Creating a table with appropriate format

```{r}

table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```


A tidy dataset must have these 3 rules:
  1. Each variable must have its own column.
  2. Each observation must have its own row.
  3. Each value must have its own cell.

## Working with the `table1` dataset #### 

The ggplot function can be used to filter, summarise and use functions such as group or golour in ggplot2, only if the data is tidy. 


```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
# mutate() is used to add a new column

```

```{r}
# Compute cases per year
table1 %>% 
  count(year, wt = cases)
# it counts the number of entries in each year
```

```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

# 2.3 Pivoting data to make it tidy

1. Need to understand what each variable and observation actually means
2. resolve the 2 common problems with untidy data:
    - one variable is spread across multiple columns
    - one observation is scattered across multiple rows
    
    
How to fix:
    - Pivot data (move it around) into tidy form
    - 2 functions are used:
        1. tidyr::pivot_longer() - makes data                longer (more rows)
        2. tidyr::pivot_wider() - makes data                  wider (more columns)

# 2.4 Lengthening datasets

pivot_longer() is used to make data longer (more rows) by gathering multiple columns into 2 columns: 
  -one for variable names and one for values.
  
For example: grouping data by year
    - use the function facet_wrap() to create a         separate plot for each year
    - use the function pivot_longer instead to split dataset by column abd reformat it into tidy format with:
        - observations as rows,
        - columns as variables
        - values as cells
        
        
        
## Using pivot_longer() in an example
```{r}
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...

```

```{r}
#using the billboard dataset for pivot_longer()

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week",    #"week","rank" are quoted because they are new variables being created
    values_to = "rank"
  )

```

## Pivoting longer - example

Using term "tribble" (not to be confused with "tibble") 
It is a type of dataframe that allows us to construct small tibbles by hand

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```

dataset = df
adding new variables:
    1. id (already exists)
    2. measurement (column names)
    3. value (cell values)
    
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
# data pivoted and repeated an id for each measurement

```

# 2.5 Widening datasets

Use the function pivot_wider() to make data wider (more columns).
It allows us to handle an observation if it is scattered across multiple rows. 

## Example of pivot_wider()

```{r}
cms_patient_experience
```

In this example, the organisation is the core unit but is spread across six rows with one row for each measure. 

The function distinct() is used to see the complete set of values of measure_cd and measure_title

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```

The columns do not have good variable names
 - measure_cd doesn't explain the meaning of variable
 - measure_title is too long to be used as a variable name
 
Measure_cd will be used as the source of new column names for this example

Pivot_wider() uses the opposite interface of pivot_longer().

We need to provide the existing column names that define the values (values_from) and column name (names_from). We do not give new column names.

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

# still not good since there are different rows for each organisation
# we need to tell pivot_wider which column have values that uniquely identify each row 
# here the variables start with "org"
```

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

## Example 2 for pivot_wider()

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

Take the names from the measurement column using the names_from() argument and the values from the value column using the values_from() argument:

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

Starting the pivoting process 
- Pivot_wider() needs to figure out what goes into rows and columns
- new column names will be unique values of measurement

```{r}
 df |> 
  distinct(measurement) |> 
  pull()

```

By default:
- rows in output are determined by all the variables without new names and values.
- these are called id_cols
- there will be only one column for this example, but normally, it can be any number of columns

```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()

```

pivot_wider() then combines these results to generate an empty dataframe:

- fills all missing values using data in input
- for this e.g. not every output has a corresponding value in the input due to no third blood pressure measurement for patient B
- cell remains empty


```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)

```


# 2.6 Pivoting exercises using palmer penguins

## Lengthening data wth pivot_longer()

- Creating a single plot showing distribution of all morphometric measurements (bill length, bill depth, flipper length) for each species of penguin
- in the current tidy format the variables are spread across four separate columns
- they can be plotted easily using Facet_wrap()
- the variables need to be stacked into 2 columns: 1. identifying measurement name
2. holding actual numerical value


Putting the columns into 'longer' format

```{r}
library(palmerpenguins)

# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)

```

structure of table after pivot_longer() has different syntax:
-  target columns we wanted to pivot (cols)
- he name of the new categorical column that will store our old column headers (names_to)
-  name of the new numeric column that will house the data (values_to).

the lengthened format is excatly what ggplot expects for multi-panel faceting

Visualising distributions simultaneously:

```{r}
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```


## Widening data with pivot_wider()

Presenting summary stats as a cross-tabulated matrix
-much easier for reader to understand a wide table instead of long, repetitive list


Calculating mean body mass for each species on each island using group_by() and summarise()

```{r}
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

```


Output above is not ideal for publication presentation, even though it is technically 'tidy' and computationally useful
- Use pivot_wider() to pull island names into columns to create a clean matrix of species by island

```{r}
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)

```

From results:
- names_from: which column provides new column headers(the islands)
- values_from:  which column populates the cells beneath them (the mean mass)
-  NA values automatically populate where a specific species does not occur on a particular island
This can be visualised directly through data wrangling


# 2.7 Separating and using data tables

Using table 3 (below)

-one column (rate) contains 2 variables (cases and population) that are separated by a "/"
- need to separate the 2 variables into their own columns using the function separate()

```{r}
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```


Splitting the rate column into 2 variables (cases and population) using the function separate()
- separate() takes the column to split, the names of the new columns, and the separator character ("/") as arguments

- by default, separate() will split values wherever it sees a non-alphanumeric character (i.e. a character that isn’t a number or letter).
-separate() split the values of rate at the forward slash characters


```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))

```


- to use a specific character to separate a column, you can pass the character to the sep argument of separate(). For example, we could rewrite the code above as:

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```

From the tables above, data types:
- Both cases and population are listed as character (<chr>) types
- This is a default of using separate()

However, since the values in those columns are actually numbers, separate() can convert them to better types using convert = TRUE

- they will be listed as integer (<int>) types instead of character (<chr>) types below:

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

```

Can also use separate() to split vactor of integers
- will interpret integers as positions to split at
-  Positive values start at 1 on the far-left of the strings
- Negative values start at -1 on the far-right of the strings

When using integers to separate strings:
-  the length of separation should be one less than the number of names in into.

Using the arrangement to separate the last two digits of each year:
- makes data less tidy, but demonstrates the flexibility of separate()

```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

```


## Using unite()

- Inverse of separate() 
- unite() combines multiple columns into a single column, with a specified separator character between the values.

Using table 5 (below) to demonstrate unite()

- used to combine multiple columns into single column
- e.g. use unite() to rejoin century and year columns
- unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select()

Note: 
Here we need to add sep ="" because we don’t want any separator (the default is to add an underscore _)


```{r}
table5 %>% 
  unite(new, century, year, sep = "")

```



# 2.8 Wrangling strings and dates

Before we can merge datasets or plot timelines, we must standardize our text strings and date-times

The tidyverse includes two specialized packages for this: 
- stringr for text 
- lubridate for dates

## Standardisong text with stringr

- All functions in the stringr package begin with str_

Cleaning a messy vector of site names before conducting a data summary

-By applying these three simple functions, we have converted five uniquely messy inputs into just two perfectly standardized categories: "nelly_bay" and "geoffrey_bay"

```{r}
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean)
  )

print(clean_sites)

```


## Mastering time with lubridate

- Allows  you to parse dates by simply spelling out the order of the components: year, month, day, hour, minute, and second.

If your dataset contains a date formatted as "25/12/2026":
- the day comes first, then the month, then the year. 
- use the dmy() function

If the date is "2026-12-25":
- use ymd().

```{r}
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

```

For high-frequency marine data, such as acoustic telemetry or CTD profiles:
- need to parse exact timestamps.
- can be done by appending _hms (hours, minutes, seconds) to our functions.


- Once a column is a recognised datetime object, ggplot2 will automatically understand how to format the x-axis chronologically
- can easily extract specific components (e.g., using month(true_time) to get the month integer).



```{r}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)

```

# 2.9 Relational data: Joining tables














