---
title: "Module 2 Workshop 2"
author: "Brihatee Mattapullut"
date: 30/09/2026 # Automatically uses today's date
date-format: "DD/MM/YYYY"             # British/Australian date format

# Output Config
format: 
  # HTML Output
  html:
    theme: cosmo          # Modern Bootstrap theme with clean typography
    highlight-style: a11y # High-contrast syntax highlighting for code
    toc: true             # Includes a Table of Contents
    toc-depth: 3          # Captures down to sub-subsections (e.g., 3.1.1)
    toc-location: left    # Pins TOC to left sidebar so it doesn't block main page
    toc-title: "Contents" # Custom title for the TOC sidebar
    number-sections: true # Numbers sections automatically (e.g., 1, 1.1, 2)
    code-fold: true       # Collapses code blocks by default under a dropdown
    code-summary: "Show code" # Label text on collapsed code blocks
    code-tools: true      # Adds global code menu (Show/Hide All Code) at top right
    code-copy: true       # Adds a one-click copy button to code blocks
    embed-resources: true # Bundles images/CSS into a single stand-alone HTML file
    df-print: paged       # Formats data frames into interactive paged tables
    fig-height: 5         # Default plot height in inches
    fig-width: 8          # Default plot width in inches
    #fig-responsive: true # Scales figures smoothly across screen sizes
    smooth-scroll: true   # Enables smooth scrolling when clicking TOC links
---

# 2.2 Tidying data - Loading libraries
```{r}
library(tidyverse)
```


Creating a table with appropriate format

```{r}

table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```


A tidy dataset must have these 3 rules:
  1. Each variable must have its own column.
  2. Each observation must have its own row.
  3. Each value must have its own cell.

## Working with the `table1` dataset #### 

The ggplot function can be used to filter, summarise and use functions such as group or golour in ggplot2, only if the data is tidy. 


```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
# mutate() is used to add a new column

```

```{r}
# Compute cases per year
table1 %>% 
  count(year, wt = cases)
# it counts the number of entries in each year
```

```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

# 2.3 Pivoting data to make it tidy

1. Need to understand what each variable and observation actually means