---
title: "Module 2 Workshop 2"
author: "Brihatee Mattapullut"
date: 30/09/2026 # Automatically uses today's date
date-format: "DD/MM/YYYY"             # British/Australian date format

# Output Config
format: 
  # HTML Output
  html:
    theme: cosmo          # Modern Bootstrap theme with clean typography
    highlight-style: a11y # High-contrast syntax highlighting for code
    toc: true             # Includes a Table of Contents
    toc-depth: 3          # Captures down to sub-subsections (e.g., 3.1.1)
    toc-location: left    # Pins TOC to left sidebar so it doesn't block main page
    toc-title: "Contents" # Custom title for the TOC sidebar
    number-sections: true # Numbers sections automatically (e.g., 1, 1.1, 2)
    code-fold: true       # Collapses code blocks by default under a dropdown
    code-summary: "Show code" # Label text on collapsed code blocks
    code-tools: true      # Adds global code menu (Show/Hide All Code) at top right
    code-copy: true       # Adds a one-click copy button to code blocks
    embed-resources: true # Bundles images/CSS into a single stand-alone HTML file
    df-print: paged       # Formats data frames into interactive paged tables
    fig-height: 5         # Default plot height in inches
    fig-width: 8          # Default plot width in inches
    #fig-responsive: true # Scales figures smoothly across screen sizes
    smooth-scroll: true   # Enables smooth scrolling when clicking TOC links
---

# 2.2 Tidying data - Loading libraries
```{r}
library(tidyverse)
```


Creating a table with appropriate format

```{r}

table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```


A tidy dataset must have these 3 rules:
  1. Each variable must have its own column.
  2. Each observation must have its own row.
  3. Each value must have its own cell.

## Working with the `table1` dataset #### 

The ggplot function can be used to filter, summarise and use functions such as group or golour in ggplot2, only if the data is tidy. 


```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
# mutate() is used to add a new column

```

```{r}
# Compute cases per year
table1 %>% 
  count(year, wt = cases)
# it counts the number of entries in each year
```

```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

# 2.3 Pivoting data to make it tidy

1. Need to understand what each variable and observation actually means
2. resolve the 2 common problems with untidy data:
    - one variable is spread across multiple columns
    - one observation is scattered across multiple rows
    
    
How to fix:
    - Pivot data (move it around) into tidy form
    - 2 functions are used:
        1. tidyr::pivot_longer() - makes data                longer (more rows)
        2. tidyr::pivot_wider() - makes data                  wider (more columns)

# 2.4 Lengthening datasets

pivot_longer() is used to make data longer (more rows) by gathering multiple columns into 2 columns: 
  -one for variable names and one for values.
  
For example: grouping data by year
    - use the function facet_wrap() to create a         separate plot for each year
    - use the function pivot_longer instead to split dataset by column abd reformat it into tidy format with:
        - observations as rows,
        - columns as variables
        - values as cells
        
        
        
## Using pivot_longer() in an example
```{r}
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...

```

```{r}
#using the billboard dataset for pivot_longer()

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week",    #"week","rank" are quoted because they are new variables being created
    values_to = "rank"
  )

```

## Pivoting longer - example

Using term "tribble" (not to be confused with "tibble") 
It is a type of dataframe that allows us to construct small tibbles by hand

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```

dataset = df
adding new variables:
    1. id (already exists)
    2. measurement (column names)
    3. value (cell values)
    
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
# data pivoted and repeated an id for each measurement

```

# 2.5 Widening datasets

Use the function pivot_wider() to make data wider (more columns).
It allows us to handle an observation if it is scattered across multiple rows. 

## Example of pivot_wider()

```{r}
cms_patient_experience
```

In this example, the organisation is the core unit but is spread across six rows with one row for each measure. 

The function distinct() is used to see the complete set of values of measure_cd and measure_title

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```

The columns do not have good variable names
 - measure_cd doesn't explain the meaning of variable
 - measure_title is too long to be used as a variable name
 
Measure_cd will be used as the source of new column names for this example

Pivot_wider() uses the opposite interface of pivot_longer().

We need to provide the existing column names that define the values (values_from) and column name (names_from). We do not give new column names.

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

# still not good since there are different rows for each organisation
# we need to tell pivot_wider which column have values that uniquely identify each row 
# here the variables start with "org"
```

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

## Example 2 for pivot_wider()

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

Take the names from the measurement column using the names_from() argument and the values from the value column using the values_from() argument:

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

